A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-11 doc/popl2013/evaluation.tex
speel cheeking

\subsubsection{Adaptive Mesh Refinement}
\label{subsec:amr}
The third application is a Legion port of an adaptive mesh refinement (AMR) benchmark 
from BoxLib \cite{BoxLib}.  The algorithm solves the two
dimensional heat diffusion equation on a grid of cells using three levels of refinement with subrefinements
randomly placed on the surface.  
%For each time step in the application, cells at the boundary of
%a refinement are interpolated from cells at a coarser level of refinement, energy fluxes between
%cells are computed, energy is transferred, and cells at a coarser level of refinement are restricted
%to the values of refined cells.  
Every levelof refinement uses a separate region, which is partitioned several ways to support multiple views of
the cells.  One partitioning separates cells into pieces that can be updated in
parallel.  Additional partitions are created for viewing data from coarser and finer levels of
the simulation.  Two types of region relationships are created: one for describing the pieces at 
each level of refinement, and another for describing the relationship between pieces at different
levels of refinement.  
%These region relationships capture both intra- and inter-level disjointness information.  
The dynamic nature of AMR requires that regions be created and partitioned at runtime.  
%The many ways in which cells are used also requires multiple simultaneous
%partitions of regions.

>>>>

\subsubsection{Adaptive Mesh Refinement}
\label{subsec:amr}
The third application is a Legion port of an adaptive mesh refinement (AMR) benchmark 
from BoxLib \cite{BoxLib}.  The algorithm solves the two
dimensional heat diffusion equation on a grid of cells using three levels of refinement with sub-refinements
randomly placed on the surface.  
%For each time step in the application, cells at the boundary of
%a refinement are interpolated from cells at a coarser level of refinement, energy fluxes between
%cells are computed, energy is transferred, and cells at a coarser level of refinement are restricted
%to the values of refined cells.  
Every level of refinement uses a separate region, which is partitioned several ways to support multiple views of
the cells.  One partitioning separates cells into pieces that can be updated in
parallel.  Additional partitions are created for viewing data from coarser and finer levels of
the simulation.  Two types of region relationships are created: one for describing the pieces at 
each level of refinement, and another for describing the relationship between pieces at different
levels of refinement.  
%These region relationships capture both intra- and inter-level disjointness information.  
The dynamic nature of AMR requires that regions be created and partitioned at runtime.  
%The many ways in which cells are used also requires multiple simultaneous
%partitions of regions.

### 2012-07-11 doc/popl2013/intro.tex
speel cheeking

In the past decade machine architecture, particularly at the high
performance end of the spectrum, has undergone a revolution.  The
latest supercomputers are now complex hierarchies of many different
kinds of computing technologies: networks of nodes at the top level,
multiple chips per node, mutiple cores within a chip, and, most
recently, multiple accelerators (usually GPUs) per node, which 
can themselves be decomposed.  We present the operational and static
semantics of Legion\cite{Legion12}, a programming model targeted at providing an
appropriate level of abstraction for programming such machines, one
that is both sufficiently high-level to be portable while still
exposing aspects that are crucial to performance.

In Legion data is organized in a hierarchy of {\em regions}
and subregions while computation is organized in a hierarchy of {\em
tasks} and subtasks operating on regions.  Regions and tasks interact
through a static system of {\em region privileges} specifying 
operations a task may perform on a region argument (read,
write, or reduce) and  {\em region coherence} annotations that
express what other tasks may do concurrently with
the region (exclusive, atomic, or simultaneous).  We prove the
soundness of the privileges and coherence system and use these two
results to prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em non-interfering} (have no ordering requirements for correctness), 
then the any descendent of $t_1$ in the task hierarchy is non-interfering
with any descendent of $t_2$.  This theorem is the basis for correct distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the huge latencies involved in communicating in the target
class of machines.

@@

Our own experience with writing high-performance applications in both 
Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple simultaneous partitions of
the same data, providing different views on to that data. Thus, in Legion the 
handling of regions and partitioning of regions is much more dynamic than in 
previous designs.  In particular,
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

% Putting this here so that we can get the code in a good place
\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={float,head,min,max}
}
\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuilt Simulation}]
--                        <voltage,current,charge,capacitance>
type CircuitNode        = <float,float,float,float>
--                     < owned node, owned or ghost node, resistance, current>
type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>

@@

The important difference bewteen Legion and the more static approaches
is that Legion allows aliasing, but does so in a structured way that minimizes
runtime overheads.  The key observation is that
static alias analysis is hard, but dynamic analysis is very easy
and relatively cheap
when done at the granularity of regions instead of individual heap
locations.  Thus, Legion falls between fully static systems, such as
Sequoia and DPJ, and fully dynamic approaches such as transactional
memory.  Legion still has a significant static component; in
particular, the required privileges for function calls are checked statically
and region
pointer dereferences are checked statically.  The dynamic alias tests
for independence are done at the granularity of regions and one check 
on a region
often serves to prove the independence of many subsequent uses of that region.
In contrast, because transactional memory has no mechanism
for grouping heap data into larger collections, it must test the
overlap between two sets of data on each individual memory access, which is
significantly more expensive.

@@

\item We give experimental evidence for the Legion design.  On three real-world applications, we show that 
dynamic region pointer checks would be expensive, justifying checking this aspect of the type system statically.
We also show that the cost of checking for non-interference of region privileges is low, showing that a much more expressive
and dynamic langauge is not incompatible with high performance.

>>>>

In the past decade machine architecture, particularly at the high
performance end of the spectrum, has undergone a revolution.  The
latest supercomputers are now complex hierarchies of many different
kinds of computing technologies: networks of nodes at the top level,
multiple chips per node, multiple cores within a chip, and, most
recently, multiple accelerators (usually GPUs) per node, which 
can themselves be decomposed.  We present the operational and static
semantics of Legion\cite{Legion12}, a programming model targeted at providing an
appropriate level of abstraction for programming such machines, one
that is both sufficiently high-level to be portable while still
exposing aspects that are crucial to performance.

In Legion data is organized in a hierarchy of {\em regions}
and subregions while computation is organized in a hierarchy of {\em
tasks} and subtasks operating on regions.  Regions and tasks interact
through a static system of {\em region privileges} specifying 
operations a task may perform on a region argument (read,
write, or reduce) and  {\em region coherence} annotations that
express what other tasks may do concurrently with
the region (exclusive, atomic, or simultaneous).  We prove the
soundness of the privileges and coherence system and use these two
results to prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em non-interfering} (have no ordering requirements for correctness), 
then the any descendant of $t_1$ in the task hierarchy is non-interfering
with any descendant of $t_2$.  This theorem is the basis for correct distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the huge latencies involved in communicating in the target
class of machines.

@@

Our own experience with writing high-performance applications in both 
Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data. Thus, in Legion the 
handling of regions and partitioning of regions is much more dynamic than in 
previous designs.  In particular,
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

% Putting this here so that we can get the code in a good place
\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={float,head,min,max}
}
\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuit Simulation}]
--                        <voltage,current,charge,capacitance>
type CircuitNode        = <float,float,float,float>
--                     < owned node, owned or ghost node, resistance, current>
type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>

@@

The important difference between Legion and the more static approaches
is that Legion allows aliasing, but does so in a structured way that minimizes
runtime overheads.  The key observation is that
static alias analysis is hard, but dynamic analysis is very easy
and relatively cheap
when done at the granularity of regions instead of individual heap
locations.  Thus, Legion falls between fully static systems, such as
Sequoia and DPJ, and fully dynamic approaches such as transactional
memory.  Legion still has a significant static component; in
particular, the required privileges for function calls are checked statically
and region
pointer dereferences are checked statically.  The dynamic alias tests
for independence are done at the granularity of regions and one check 
on a region
often serves to prove the independence of many subsequent uses of that region.
In contrast, because transactional memory has no mechanism
for grouping heap data into larger collections, it must test the
overlap between two sets of data on each individual memory access, which is
significantly more expensive.

@@

\item We give experimental evidence for the Legion design.  On three real-world applications, we show that 
dynamic region pointer checks would be expensive, justifying checking this aspect of the type system statically.
We also show that the cost of checking for non-interference of region privileges is low, showing that a much more expressive
and dynamic language is not incompatible with high performance.
